### 2014-11-13 tex/bisimulation.tex
Finished pass on bisimulation.

\section{Bisimilarity Relation}
\label{sec:bisim}
Let $p \rgse p^\prime$ be the binary relation over the universe of states, $S$,
according to generalized symbolic execution with lazy initialization
(see the supplemental text for a precise definition of that
relation). Similarly, let $q \rsym q^\prime$ be the relation for the GSESH
algorithm in this paper. The principle theoretical result is the
existence of a bisimilarity relation between states related on $p \rgse p^\prime$
and the states related on $q \rsym q^\prime$. A corollary is that $\rsym$ is
both sound and complete with respect to $\rgse$. Only a summary of the
proof is given due to space.

@@

Let $p \rgse p^\prime$ be defined as a union over state relations:
$$
\rgse = \rgse^\mathit{A} \cup  \rgse^\mathit{A^\prime} \cup \rgse^\mathit{W} \cup \rgse^\mathit{W^\prime} \cup \rgse^\mathit{E} \cup \rgse^\mathit{E^\prime} \cup \rgse^\com
$$
where \emph{A} is a field access $\lp \cfgt{*}\ \cfgt{\$}\ \cfgnt{f}
\rightarrow \cfgnt{k}\rp$, \emph{W} is a field write $\lp
\cfgnt{x}\ \cfgt{\$}\ \cfgnt{f}\ \cfgt{:=}\ \cfgnt{*} \rightarrow
\cfgnt{k}\rp$, \emph{E} is a reference compare $\lp
\cfgnt{v}\ \cfgt{=}\ \cfgnt{*} \rightarrow \cfgnt{k}\rp$, the prime symbol
indicates a null reference in the operation or a false outcome, and
\emph{\com} is everything else in the language. Any state relation, say $\rightarrow_x$, is trivially extended to sets of states as
$$
P \hookrightarrow_x P^\prime \Longleftrightarrow \forall p \in P\ (\forall p^\prime\ (p \rightarrow_x p^\prime \Rightarrow p^\prime \in P^\prime))
$$
Let $\hookrightarrow_\gse$ be the extension of $\rgse$ to sets of states.
$$
\rsgse = \hookrightarrow_\gse^\mathit{A}
\cup \hookrightarrow_\gse^\mathit{A^\prime} \cup \hookrightarrow_\gse^\mathit{W} \cup
\hookrightarrow_\gse^\mathit{W^\prime} \cup \hookrightarrow_\gse^\mathit{E} \cup \hookrightarrow_\gse^\mathit{E^\prime}
\cup \hookrightarrow_\gse^\com
$$
The definition embodies the notion of splitting equivalence classes at
certain operations. The set $P$ at a field access for example has two
potential successors: a non-null and null reference. Respectively, the
set $P \rgse^\com P^\prime$ only has a single successor since those expressions
are not related to control flow.

The functional equivalence between heaps in states $p$
and $q$ requires both a mapping to relate the two heaps and a
constraint on the feasibility of that mapping in the presence of a
path constraint from symbolic execution. Subscripts indicates state
members as in $p = (
\cfgnt{L}_p\ \cfgnt{R}_p\ \phi_p\ \eta_p\ \cfgnt{e}_p\ \cfgnt{k}_p )$.
\begin{definition}
\label{def:homomorphism}
A \textbf{homomorphism} given the universe of field names $\mathcal{F}$ is 
$$
\begin{array}{l}
 s_p \rightharpoonup_{h} s_q \Leftrightarrow \\
\ \ \ \ \exists h: \mathcal{L} \mapsto \mathcal{L}\ (\forall \cfgnt{l}_\alpha\ (\forall \cfgnt{l}_\beta\ (\forall f \in \mathcal{F}\ ( \exists \phi_\alpha\ (\exists \phi_\beta\ ( \\ 
\ \ \ \ \ \ \ \ \ \ \ \ (\phi_\alpha\ \cfgnt{l}_\alpha) \in \cfgnt{L}_p(\cfgnt{R}_p (\cfgnt{l}_\beta,f )) \Rightarrow \\
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (\phi_\beta\ h(\cfgnt{l}_\alpha))\in \cfgnt{L}_q(\cfgnt{R}_q (h(\cfgnt{l}_\beta),f ))\ \\
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \  )) ) ) ) )
\end{array}
$$
\end{definition}

\begin{definition}
\label{def:hc}
The \textbf{homomorphism constraint} is
\begin{align*}
\mathbb{HC}(p \rightharpoonup_{h} q) &= \\
 \bigwedge \{ \phi_b\ | \exists (\phi_a\ l) \in \cfgnt{L}_p^\rightarrow ( (\phi_b\ h(l)) \in \cfgnt{L}_q^\rightarrow)\} 
\end{align*}
\end{definition}
The functional equivalence between heaps on a homomorphism and
a homomorphism constraint asserts a common structures in the two heaps under certain
conditions. The equivalence is used to related states in $p \rgse
p^\prime$ to states in $q \rsym q^\prime$.
\begin{definition}
\label{representation}
States $(p\ q)$ are in the \textbf{representation relation}, $p \sqsubset q$, if and only if, $\eta_p = \eta_q ,\ \cfgnt{e}_p =
\cfgnt{e}_q ,\ \cfgnt{k}_p = \cfgnt{k}_q$, and there exists a
homomorphism $p \rightharpoonup_{h}
q$ such that
\begin{equation}
\label{eqn:valid}
 \mathbb{S}( \phi_q \wedge \mathbb{HC}(s_p \rightharpoonup_{h} s_q) ) 
\end{equation}
The represented relation is extended to sets of states $P$ and a single state $q$ as\footnote{The full proof argues $P \sqsubset Q$ but that $Q$ is
  always a singleton set containing $q$.}
$$
P \sqsubset q \Longleftrightarrow \forall p\ (p \sqsubset q \Rightarrow p \in P)
$$
\end{definition}
The statement $p \sqsubset q$ ensures that a functionally equivalent
heap to the one in $p$ is present, by the homomorphism, and valid, by
the heap constraint and path constraint, in $q$. As the states in $P$
are only differentiated by heaps and those states are only
differentiated from $q$ by both the heap and path constraint in $q$,
$P \sqsubset q$ implies that $q$ is representative of all the states
in $P$ up to the given point of execution expressed in $\phi_q$.
\begin{definition}
\label{bisimulation}
The \textbf{functional associated to bisimulation} applied to $\sqsubset$, denoted as $F_\sim(\sqsubset)$, is the set of all pairs
$(P\ q)$ such that
\begin{equation}
\label{eqn:BisimulationForwards}
\forall P^\prime\ ( P \rsgse P^\prime \Rightarrow \exists q^\prime\ ( q \rsym q^\prime \wedge P^\prime\ \sqsubset\ q^\prime))
\end{equation}
\begin{equation}
\label{eqn:BisimulationBackwards}
\forall q^\prime\ ( q \rsym q^\prime\Rightarrow \exists P^\prime\ ( P \rsgse P^\prime \wedge P^\prime\ \sqsubset\ q^\prime))
\end{equation}
If $\sqsubset$ is a bisimulation, then the greatest fixed point of $F_\sim(\sqsubset)$ is the bisimilarity relation denoted by $\sim$.
\end{definition}
As is typical, the functional reasons over a forward and backward
simulation, while the application to dissimilar states using a
meta-relation, $\rsgse$, is somewhat non-standard; though convenient
in this instance since it captures the notion of $q$ as a set of
states \cite{GSE:barbedbisimulation}.

The state $S$ is the universe of well-formed states, and the
set $S_A \subseteq S$ is the set of states at a field access continuation.
\begin{lemma}[\textrm{F{\footnotesize IELD}} \textrm{A{\footnotesize CCESS}} preserves $\sqsubset\ \subseteq F_\sim(\sqsubset)$]
If $P \in 2^{S_\mathit{FA}}$ and $q \in S$ are such that $P \sqsubset q$, then $(P\ q)$ is in the functional associated to bisumlation.
\label{lem:access}
$$
\forall P \in 2^{S_\mathit{FA}}\ (P \sqsubset q \Rightarrow (P\ q) \in F_\sim(\sqsubset))
$$
\end{lemma}

@@

In the backward simulation \eqref{eqn:BisimulationBackwards}, for each
$q^\prime$ such that $q \rsym q^\prime$, the proof posits the
existences of some $p^\prime \sqsubset q^\prime$ and uses the
homomorphism in $p^\prime \sqsubset q^\prime$ to derive a $p$ such
that $p \rgse p^\prime$ and $p \in P$.

As the forward and backward simulations hold for any $p \in P$, $P
\sqsubset q$ must be in the functional, $F_\sim(\sqsubset)$, which is
a contradiction.
\end{proof}
Similar lemmas are proved for field write and equals reference. These
require additional lemmas on properties of the summary machine
$\rightarrow_S$. Specifically, that it preserves determinism, and it
preserves the homomorphism and the satisfiability of homomorphism constraint

\begin{theorem}
\label{th:bisim}
The relation $\sqsubset$ is a bisimulation: $\sqsubset\ \subseteq\ \sim$
\end{theorem}
\begin{proof}
By definition of the bisimilarity relation. All of the common rules
make no changes to the heap, so $P \sqsubset q$ is included in the
functional. The other transition relations are supported by lemmas
such as \lemref{lem:access}.
\end{proof}

The notation $p \stackrel{n}{\rgse} p^\prime$ denotes that $p^\prime$
is the $n$-step successor of $p$, if it exists, and $q
\stackrel{n}{\rsym} q^\prime$ is similarly defined. Completeness and
soundness are readily established by \thref{th:bisim} and inducting
over $n$.

>>>>

\section{Bisimilarity Relation}
\label{sec:bisim}
Recall that $p \rgse p^\prime$ is the binary relation over the universe of states, $S$,
according to GSE with lazy initialization. Similarly, $q \rsym q^\prime$ is the relation for the summary heap algorithm
presented in this paper. The principle theoretical result is the
existence of a bisimulation between states related on $p \rgse p^\prime$
and the states related on $q \rsym q^\prime$.  A corollary is that $\rsym$ is
both sound and complete with respect to properties provable in $\rgse$. Only a summary of the
proofs is given due to space.

@@

Let $p \rgse p^\prime$ be defined as a union over state relations:
$$
\rgse = \rgse^\mathit{A} \cup  \rgse^\mathit{A^\prime} \cup \rgse^\mathit{W} \cup \rgse^\mathit{W^\prime} \cup \rgse^\mathit{E} \cup \rgse^\mathit{E^\prime} \cup \rcom
$$
where \emph{A} is a field access after evaluating the expression for the base reference $\lp \cfgt{*}\ \cfgt{\$}\ \cfgnt{f}
\rightarrow \cfgnt{k}\rp$, \emph{W} is a field write after evaluating the expression for the right operand$\lp
\cfgnt{x}\ \cfgt{\$}\ \cfgnt{f}\ \cfgt{:=}\ \cfgnt{*} \rightarrow
\cfgnt{k}\rp$, \emph{E} is a reference compare after evaluating the left and right operand expressions $\lp
\cfgnt{v}\ \cfgt{=}\ \cfgnt{*} \rightarrow \cfgnt{k}\rp$, the prime symbol
indicates a null reference in the operation or a false outcome, and
\emph{\com} is everything else in the language. Any state relation, say $\rightarrow_x$, is trivially extended to sets of states as
$$
P \hookrightarrow_x P^\prime \Longleftrightarrow \forall p \in P\ (\forall p^\prime\ (p \rightarrow_x p^\prime \Rightarrow p^\prime \in P^\prime))
$$
Let $\hookrightarrow_\gse$ be the extension of $\rgse$ to sets of
states. From this extension, a new meta transition relation is defined
over sets of states as
$$
\rsgse = \hookrightarrow_\gse^\mathit{A}
\cup \hookrightarrow_\gse^\mathit{A^\prime} \cup \hookrightarrow_\gse^\mathit{W} \cup
\hookrightarrow_\gse^\mathit{W^\prime} \cup \hookrightarrow_\gse^\mathit{E} \cup \hookrightarrow_\gse^\mathit{E^\prime}
\cup \hookrightarrow_\com
$$
The definition embodies the notion of splitting groups of heaps at
certain operations. The set $P$ at a field access after base reference is known, for example, has two
potential successors: a non-null and null reference. The $\rsgse$ relation has two successors at that point to divide $P$ into the two outcomes. In contrast, the
set $P \rgse^\com P^\prime$ only has a single successor since anything in $\rightarrow_\com$ is not related to control flow (e.g., variable look up, creating continuations, etc.).

The functional equivalence between heaps in states $p$
and $q$ requires both a mapping to relate the two heaps and a
constraint on the feasibility of that mapping in the presence of a
path constraint from symbolic execution. Subscripts indicates state tuple
members as in $p = (
\cfgnt{L}_p\ \cfgnt{R}_p\ \phi_p\ \eta_p\ \cfgnt{e}_p\ \cfgnt{k}_p )$.
\begin{definition}
\label{def:homomorphism}
A \textbf{homomorphism} given the universe of field names $\mathcal{F}$ is 
$$
\begin{array}{l}
 s_p \rightharpoonup_{h} s_q \Leftrightarrow \\
\ \ \ \ \exists h: \mathcal{L} \mapsto \mathcal{L}\ (\forall \cfgnt{l}_\alpha\ (\forall \cfgnt{l}_\beta\ (\forall f \in \mathcal{F}\ ( \exists \phi_\alpha\ (\exists \phi_\beta\ ( \\ 
\ \ \ \ \ \ \ \ \ \ \ \ (\phi_\alpha\ \cfgnt{l}_\alpha) \in \cfgnt{L}_p(\cfgnt{R}_p (\cfgnt{l}_\beta,f )) \Rightarrow \\
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ (\phi_\beta\ h(\cfgnt{l}_\alpha))\in \cfgnt{L}_q(\cfgnt{R}_q (h(\cfgnt{l}_\beta),f ))\ \\
\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \  )) ) ) ) )
\end{array}
$$
\end{definition}

\begin{definition}
\label{def:hc}
The \textbf{homomorphism constraint} is
\begin{align*}
\mathbb{HC}(p \rightharpoonup_{h} q) &= \\
 \bigwedge \{ \phi_b\ | \exists (\phi_a\ l) \in \cfgnt{L}_p^\rightarrow ( (\phi_b\ h(l)) \in \cfgnt{L}_q^\rightarrow)\} 
\end{align*}
\end{definition}
The functional equivalence between heaps on a homomorphism and
a homomorphism constraint asserts a common structure in the two heaps under certain
conditions. The equivalence is used to related states in $p \rgse
p^\prime$ to states in $q \rsym q^\prime$.
\begin{definition}
\label{representation}
States $(p\ q)$ are in the \textbf{representation relation}, $p \sqsubset q$, if and only if, $\eta_p = \eta_q ,\ \cfgnt{e}_p =
\cfgnt{e}_q ,\ \cfgnt{k}_p = \cfgnt{k}_q$, and there exists a
homomorphism $p \rightharpoonup_{h}
q$ such that
\begin{equation}
\label{eqn:valid}
 \mathbb{S}( \phi_q \wedge \mathbb{HC}(s_p \rightharpoonup_{h} s_q) ) 
\end{equation}
The represented relation is extended to sets of states $P$ and a single state $q$ as\footnote{The full proof argues $P \sqsubset Q$ but that any member $Q$ represents that same set of heaps as any other member of $q$ so it is sufficient to a member of $q$ in the relation.}
$$
P \sqsubset q \Longleftrightarrow \forall p\ (p \sqsubset q \Rightarrow p \in P)
$$
\end{definition}
The statement $p \sqsubset q$ ensures that a functionally equivalent
heap to the one in $p$ is present, by the homomorphism, and valid, by
the heap constraint and path constraint, in $q$. As the states in $P$
are only differentiated by heaps and those states are only
differentiated from $q$ by both the heap and path constraint in $q$,
$P \sqsubset q$ implies that $q$ is representative of all the states
in $P$ up to the given point of execution expressed in $\phi_q$.
\begin{definition}
\label{bisimulation}
The \textbf{functional associated to bisimulation} applied to $\sqsubset$, denoted as $F_\sim(\sqsubset)$, is the set of all pairs
$(P\ q)$ such that
\begin{equation}
\label{eqn:BisimulationForwards}
\forall P^\prime\ ( P \rsgse P^\prime \Rightarrow \exists q^\prime\ ( q \rsym q^\prime \wedge P^\prime\ \sqsubset\ q^\prime))
\end{equation}
\begin{equation}
\label{eqn:BisimulationBackwards}
\forall q^\prime\ ( q \rsym q^\prime\Rightarrow \exists P^\prime\ ( P \rsgse P^\prime \wedge P^\prime\ \sqsubset\ q^\prime))
\end{equation}
If $\sqsubset$ is a bisimulation, then the greatest fixed point of $F_\sim(\sqsubset)$ is the bisimilarity relation denoted by $\sim$.
\end{definition}
As is typical, the functional reasons over a forward and backward
simulation, while the application to dissimilar states using a
meta-relation, $\rsgse$, is somewhat non-standard; though it is convenient
in this instance since it captures the notion of $q$ as a set of
states \cite{GSE:barbedbisimulation}.

In the following lemma, recall that $S$ is the universe of well-formed states, and the
set $S_A \subseteq S$ is the set of states at a field access continuation having computed the base reference.
\begin{lemma}[\textrm{F{\footnotesize IELD}} \textrm{A{\footnotesize CCESS}} preserves $\sqsubset\ \subseteq F_\sim(\sqsubset)$]
If $P \in 2^{S_\mathit{FA}}$ and $q \in S$ are such that $P \sqsubset q$, then $(P\ q)$ is in the functional associated to bisumlation.
\label{lem:access}
$$
\forall P \in 2^{S_\mathit{FA}}\ (P \sqsubset q \Rightarrow (P\ q) \in F_\sim(\sqsubset))
$$
\end{lemma}

@@

In the backward simulation \eqref{eqn:BisimulationBackwards}, for each
$q^\prime$ such that $q \rsym q^\prime$, the proof posits that with any 
$p^\prime$ such that $p^\prime \sqsubset q^\prime$, it is possible to use the
homomorphism in $p^\prime \sqsubset q^\prime$ to derive a $p$ such
that $p \rgse p^\prime$ and $p \in P$.

As the forward and backward simulations hold for any $p \in P$, $P
\sqsubset q$ must be in the functional, $F_\sim(\sqsubset)$, which is
a contradiction.
\end{proof}
Similar lemmas are proved for field write and equals reference. These
require additional lemmas on properties of the summarize machine
$\rightarrow_S$. Specifically, that it preserves determinism, it
preserves the homomorphism, and it preserves the satisfiability of the homomorphism constraint

\begin{theorem}
\label{th:bisim}
The relation $\sqsubset$ is a bisimulation: $\sqsubset\ \subseteq\ \sim$
\end{theorem}
\begin{proof}
By definition of bisimilarity. All of the common rules in $\rcom$
make no changes to the heap, so $P \sqsubset q$ is included in the
functional. The other rules that affect the heap are supported by lemmas
such as \lemref{lem:access}.
\end{proof}

The notation $p \stackrel{n}{\rgse} p^\prime$ denotes that $p^\prime$
is the $n$-step successor of $p$, when it exists, and $q
\stackrel{n}{\rsym} q^\prime$ is similarly defined. Completeness and
soundness are readily established by \thref{th:bisim} and inducting
over $n$.
